Tools for identifying unexpectedly low microfossil count sums

### 2019-05-19 Rmd/count_check_MS.Rmd
tweaks

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so. This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets fail these tests.
journal: "Quaternary Science Reviews"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
output:
  bookdown::pdf_book:
    base_format: 
      rticles::elsevier_article
    number_sections: true
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/journal-of-paleolimnology.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = TRUE, message = TRUE)

@@

Dataset diatom1 includes `r diatom1_summ$nsamples` diatom assemblages with a median richness of `r diatom1_summ$median_taxa` taxa.
The associated paper reports that at least 400 valves were counted from each assemblage, which means that singletons should have a relative abundance of 0.25% or less.
However, the rarest taxon in two assemblages have a relative abundance of `r diatom1_summ$maxpc`%, and the relative abundance of all other taxa in these assemblages are, within rounding error, integer multiples of this`r diatom1_summ$prop_integer %>% insist(. == 1) %>% invisible()`.
If the count sums actually are at least 400, this would imply that each species count in these assemblages is an integer multiple of at least `r diatom1_summ$integer_mult`.
This seems unlikely.
Alternatively, the actual count sum for these assemblages could be as low as `r diatom1_summ$low_n` valves.
In total, `r diatom1_summ$n_low_n %>% as.english()` assemblages appear to have a count sum below 400 valves.

>>>>

---
title: "Tools for identifying unexpectedly low microfossil count sums"
author:
  - name: Richard J. Telford
    email: richard.telford@uib.no
    affiliation: University of Bergen
    footnote: Corresponding Author
address:
  - code: University of Bergen
    address: Department of Biological Sciences, University of Bergen and Bjerknes Centre for Climate Research, Post Box 7803, N-5020 Bergen, Norway
abstract: |
  Microfossil counts are a key data type in palaeoecology. Recent work has raised the possibility that some authors might misreport an important quality control parameter, the counts sums, occasionally dramatically so. This paper introduces methods that can flag assemblages with potentially misreported count sums and finds that some assemblage datasets fail these tests.
journal: "Journal of Quaternary Science"
date: "`r format(Sys.Date(), '%d %B %Y')`"
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{pdflscape}
   - \newcommand{\blandscape}{\begin{landscape}}
   - \newcommand{\elandscape}{\end{landscape}}   
output:
  bookdown::pdf_book:
    base_format: 
      rticles::elsevier_article
    number_sections: true
documentclass: article
classoption: a4paper
bibliography: extra/countMS2.bib
csl: extra/elsevier-harvard_rjt.csl
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, include = TRUE, warning = TRUE, message = TRUE, error = TRUE)

@@

Dataset diatom1 includes `r diatom1_summ$nsamples` diatom assemblages with a median richness of `r diatom1_summ$median_taxa` taxa.
The associated paper reports that at least 400 valves were counted from each assemblage, which means that singletons should have a relative abundance of 0.25% or less.
However, the rarest taxon in two assemblages have a relative abundance of `r diatom1_summ$maxpc`%, and the relative abundance of all other taxa in these assemblages are, within rounding error, integer multiples of this`r diatom1_summ$prop_integer %>% insist(. == 1) %>% invisible()`.
If the count sums actually are at least 400, this would imply that each species count in these assemblages is an integer multiple of at least `r diatom1_summ$integer_mult`.
This seems unlikely.
Alternatively, the actual count sum for these assemblages could be as low as `r diatom1_summ$low_n` valves.
In total, `r diatom1_summ$n_low_n %>% as.english()` assemblages appear to have a count sum below 400 valves. 
The results section of the paper acknowledges that counts were below the standard 400 valves for part of the Late Glacial.
All the assemblages identified with possible low count sums are in this part of the stratigraphy, demonstrating the utility of the method.
